---
title: "Top Five Teams By WP"
author: "Data Punk"
date: "2025-09-02"
output: html_document
---

License Information

MIT License

Copyright (c) 2024 Data Punk Media

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

 
# Load libraries for the script.

```{r}

library(tidyverse)
library(dplyr)
library(rvest)
library(xml2)

```

# Scrape the data from Hockey-Reference.com. 

```{r}

url <- "https://www.hockey-reference.com/playoffs/NHL_2025.html"

team_statistics <- url %>% 
  read_html() %>% 
  html_nodes(xpath = '//*[@id="teams"]') %>% 
  html_table()

winprob_statistics <- team_statistics[[1]]

team_pwp_df <- as.data.frame(winprob_statistics)

team_pwp_df <- team_pwp_df %>% 
  filter(Team != "League Average")


colnames(team_pwp_df) <- c("RANK", "TEAM", "GP", "W", "L", "T", "OW", "OL",
                              "WL_PCT", "GF", "GA", "DIFF")

print(team_pwp_df)

```

# Clean up the data and transform it. 

```{r}

team_abbr_lookup <- data.frame(
  TEAM = c(
    "Anaheim Ducks", "Arizona Coyotes", "Boston Bruins", "Buffalo Sabres",
    "Calgary Flames", "Carolina Hurricanes", "Chicago Blackhawks", "Colorado Avalanche",
    "Columbus Blue Jackets", "Dallas Stars", "Detroit Red Wings", "Edmonton Oilers",
    "Florida Panthers", "Los Angeles Kings", "Minnesota Wild", "Montreal Canadiens",
    "Nashville Predators", "New Jersey Devils", "New York Islanders", "New York Rangers",
    "Ottawa Senators", "Philadelphia Flyers", "Pittsburgh Penguins", "San Jose Sharks",
    "Seattle Kraken", "St. Louis Blues", "Tampa Bay Lightning", "Toronto Maple Leafs",
    "Utah Hockey Club", "Vancouver Canucks", "Vegas Golden Knights", "Washington Capitals",
    "Winnipeg Jets"
  ),
  TEAM_ABB = c(
    "ANA", "ARI", "BOS", "BUF",
    "CGY", "CAR", "CHI", "COL",
    "CBJ", "DAL", "DET", "EDM",
    "FLA", "LAK", "MIN", "MTL",
    "NSH", "NJD", "NYI", "NYR",
    "OTT", "PHI", "PIT", "SJS",
    "SEA", "STL", "TBL", "TOR",
    "UTA", "VAN", "VGK", "WSH",
    "WPG"
  ),
  stringsAsFactors = FALSE
)

team_pwp_df <- team_pwp_df %>%
  left_join(team_abbr_lookup, by = "TEAM")

team_pwp_df$SEASON = "2024-2025"

team_pwp_df$PWP = (team_pwp_df$GF^2)/(team_pwp_df$GF^2+team_pwp_df$GA^2)

team_pwp_df <- team_pwp_df %>% 
  select(SEASON, RANK, TEAM, TEAM_ABB, GP, W, L, T, OW, OL, WL_PCT, GF, GA,
         DIFF, PWP)

print(team_pwp_df)

write.csv(team_pwp_df, "nhl_team_by_pwp.csv", row.names = FALSE)

```

# Trim the data to a smaller dataset.

```{r}

team_summary <- team_pwp_df %>% 
  select(SEASON, TEAM_ABB, GF, GA, PWP)


team_summary$PWP <- round(team_summary$PWP * 100, 2)

print(team_summary)

write.csv(team_summary, "trimmed_team_summary.csv", row.names = FALSE)

```


```{r}

cuts <- quantile(team_summary$PWP, probs = c(0, .25, .5, .75, 1), na.rm = TRUE, type = 7)
cuts

```


```{r}

team_summary <- team_summary %>%
  mutate(BRACKET = cut(
    PWP,
    breaks = unique(cuts),
    include.lowest = TRUE,
    labels = c("4","3","2","1")
  ))

team_summary

write.csv(team_summary, "team_summary_w_quartiles.csv", row.names = FALSE)

```

```{r}

filter_teams <- function(data, teams) {
  data %>%
    filter(TEAM_ABB %in% teams) %>%
    select(TEAM_ABB, PWP)
}

result <- filter_teams(team_summary, c("EDM", "TOR"))


write.csv(result, "matchup_results.csv", row.names = FALSE)

print(result)

```

# Create a Head-to-Head prediction function. 

```{r}

head_to_head_prediction <- function(team1, team2, season) {
  df1 <- team_summary %>% filter(TEAM_ABB == team1, SEASON == season)
  df2 <- team_summary %>% filter(TEAM_ABB == team2, SEASON == season)
  
  if (nrow(df1) == 0 || nrow(df2) == 0) {
    return("One or both teams not found.")
  }
  
  pwp1 <- df1$PWP
  pwp2 <- df2$PWP
  
  prob_team1_win <- pwp1 / (pwp1 + pwp2)
  prob_team2_win <- pwp2 / (pwp1 + pwp2)
  
  cat("\n=== Head-to-Head Prediction ===\n")
  cat("Season:", season, "\n")
  cat(team1, "Win Probability:", sprintf("%.1f%%", 100 * prob_team1_win), "\n")
  cat(team2, "Win Probability:", sprintf("%.1f%%", 100 * prob_team2_win), "\n")
}

```

# Print out the head-to-head using 2 teams as input. 

```{r}

head_to_head_prediction("EDM", "TOR", "2024-2025")

```










